---
title: "Austin Bikeshare Analysis"
author: Daryll
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
---



## Goals

- Identify the optimal time window for performing station and bicycle maintenance that minimizes disruption to bikeshare operations, based on usage patterns across different hours of the day.
- Uncover and present additional meaningful insights from the dataset that could support operational decisions or improve the overall efficiency of the bikeshare system.

```{r setup, message=FALSE}
# load libraries
library(tidyverse)
library(lubridate)
library(janitor)
library(ggplot2)
library(scales)
```

```{r load-data}
# Load datasets
bikeshare_df <- read_csv("bikeshare_trips.csv")
station_df <- read_csv("bikeshare_station.csv")

# Clean column names
bikeshare_df <- clean_names(bikeshare_df)
station_df <- clean_names(station_df)
```

## Data Overview

```{r data-preview}
glimpse(bikeshare_df)
glimpse(station_df)
```

## Convert and Parse Timestamps

```{r date-time-formatting}
bikeshare_df <- bikeshare_df %>%
    mutate(
        start_time = ymd_hms(start_time),
        duration_minutes = as.numeric(start_time, units = "mins")
    )
```

## Outlier Handling (Duration)

```{r outlier-capping}
# Use IQR method to cap outliers
q1 <- quantile(bikeshare_df$duration_minutes, 0.25, na.rm = TRUE)
q3 <- quantile(bikeshare_df$duration_minutes, 0.75, na.rm = TRUE)
iqr <- q3 - q1

lower_bound <- q1 - 1.5 * iqr
upper_bound <- q3 + 1.5 * iqr

bikeshare_df <- bikeshare_df %>%
    mutate(duration_capped = pmin(pmax(duration_minutes, lower_bound), upper_bound))

# Add a boxplot to confirm outlier handling
ggplot(bikeshare_df, aes(x = "", y = duration_capped)) +
    geom_boxplot(fill = "lightblue") +
    labs(title = "Capped Trip Duration Distribution", y = "Duration (minutes)")
```

## Rides by Hour of Day

```{r hourly-trips}
bikeshare_df %>%
    mutate(hour = hour(start_time)) %>%
    count(hour) %>%
    ggplot(aes(x = hour, y = n)) +
    geom_col(fill = "steelblue") +
    scale_y_continuous(labels = comma) +
    labs(title = "Bikeshare Trips by Hour of Day", x = "Hour", y = "Number of Trips")
```

## Monthly Trend

```{r monthly-trips}
bikeshare_df %>%
    mutate(year_month = floor_date(start_time, "month")) %>%
    count(year_month) %>%
    ggplot(aes(x = year_month, y = n)) +
    geom_line(color = "darkgreen") +
    labs(title = "Monthly Bikeshare Usage Over Time", x = "Month", y = "Trips")
```

## Map of Stations by Status
```{r station-map}
library(sf)
library(ggplot2)
library(maps)
# Extract latitude and longitude from 'location' column if present
if ("location" %in% names(station_df)) {
  station_df <- station_df %>%
    mutate(
      location = str_remove_all(location, "[()]"),
      latitude = as.numeric(str_split_fixed(location, ",\\s*", 2)[,1]),
      longitude = as.numeric(str_split_fixed(location, ",\\s*", 2)[,2])
    )
  # Use extracted columns for mapping
  station_df$lat <- station_df$latitude
  station_df$lon <- station_df$longitude
}

# Convert station_df to sf object
station_sf <- station_df %>%
  filter(!is.na(lat), !is.na(lon)) %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

# Get Texas map for context (Austin is in Texas)
texas_map <- map_data("state") %>% filter(region == "texas")


bbox <- st_bbox(station_sf)

xlim_buffered <- bbox[c("xmin", "xmax")] + c(-0.05, 0.05)
ylim_buffered <- bbox[c("ymin", "ymax")] + c(-0.05, 0.05)

# Plot stations colored by status
ggplot() +
  geom_polygon(
    data = texas_map,
    aes(x = long, y = lat, group = group),
    fill = "gray95", color = "gray80"
  ) +
  geom_sf(
    data = station_sf,
    aes(color = status),
    size = 2, alpha = 0.8
  ) +
  scale_color_manual(values = c("open" = "green", "closed" = "red")) +
  
  # --- AMENDED LINE USING THE AUTOMATIC BOUNDS ---
  coord_sf(xlim = xlim_buffered, ylim = ylim_buffered) +
  
  labs(
    title = "Austin Bikeshare Stations by Status",
    color = "Station Status"
  ) +
  theme_minimal()
```

## Conclusion

- Stakeholders are looking for the best time to service their station and bicycles with minimal interruption on the bikeshare service between 12am–6am.
- Sharebike shows a consistent growth in usage over time, but there are two spikes during year 2018 which need further investigation. while a drop in usage during 2020 may be related to the COVID-19 pandemic.
- Station closure status appears random; more investigation into station metadata or operational logs is recommended.

## Next Step:

- Investigate and correlate the 2018 usage spikes and the 2020 drop with external events and data.
- Perform anomaly detection or clustering on closed stations to find non-obvious patterns.
- Simulate downtime impacts during early morning hours to quantify disruption for maintenance planning.
